What We Eval in the Shadows

### 2021-04-12 analysis/environments.Rmd
Environments: table for new in progress

```{r envir_most_common}
frequent_envir <- E_envir %>% group_by(src_ref) %>% 
  summarize(n_envir = n_distinct(envir), envir = paste0(unique(envir), collapse = " ; "), envir_type = paste0(unique(envir_type), collapse = " ; "), n_types = n_distinct(envir_type))

@@

# Caller environments

# New environments

@@

>>>>

```{r envir_most_common}
frequent_envir <- E_envir %>% group_by(src_ref) %>% 
  summarize(n_envir = n_distinct(envir), envir = paste0(sort(unique(envir)), collapse = " ; "), envir_type = paste0(unique(envir_type), collapse = " ; "), n_types = n_distinct(envir_type))

@@

@@

# Less fine-grained classification

We categorize environments in:

- function environments (parent frames and integer)
- package/namespace environments
- list/NULL environments
- global environment
- new environment
- empty environment

## Caller environments

## New environments

We classify the most specific environment or caller environment after it.
```{r}
env_classes %>% filter(!is.na(new_env)) %>% group_by(env_class) %>% 
  summarise(n = n_distinct(src_ref)) %>% 
  arrange(desc(n)) %>% 
  mutate(pc = ratio(n, nb_call_sites), cpc = cumsum(pc))
```

And with the higher-level categories:

```{r}
high_level_new <- env_calls %>% filter(!is.na(new_env)) %>% 
  mutate(category = case_when(
    !is.na(hierarchy) ~ "function",
    specific_env == "global" ~ "global",
    specific_env == "empty" ~ "empty",
    str_detect(specific_env, fixed("loop")) ~ "other",
    !is.na(specific_env) ~ "package"
  ))

high_level_new_sites <- high_level_new %>%group_by(category) %>%
  summarise(n = n_distinct(src_ref)) %>% 
  arrange(desc(n)) %>% 
  mutate(pc = ratio(n, nb_new_env_sites), cpc = cumsum(pc))

nrows <- 5
r_vec("new env category name", nrows, high_level_new_sites$category)
r_vec("new env category sites", nrows, high_level_new_sites$n)
r_vec("new env category site percent", nrows, high_level_new_sites$pc)
```

## The final high-lelve classification

```{r}
nb_dataframe_env_sites <- nb_env_list_sites + nb_null_sites

nb_pack_sites <- nb_package_env_sites + nb_namespace_env_sites + nb_base_env_sites

nb_function_env_sites <- nb_caller_env_sites + nb_integer_sites

overview_table(
  r("nb dataframe env sites", nb_dataframe_env_sites),
  r("nb dataframe env site percent", ratio(nb_dataframe_env_sites, nb_call_sites)),
  r("nb package namespace env sites", nb_pack_sites),
  r("nb package namespace env site percent", ratio(nb_pack_sites, nb_call_sites)),
  r("nb function env sites", nb_function_env_sites),
  r("nb function env site percent", ratio(nb_function_env_sites, nb_call_sites))
)
```

### 2021-04-12 paper/main.tex
Environments: table for new in progress

The top-level environment in R is called the global environment. New
environments can be created using \c{new.env}, or \c{list2env}. They can be
provided a parent environment which becomes the enclosing scope of the new
environment.
% as.environment also can be used sometimes for that
% list2env creates an environment only if its argument envir is not provided

Another kind of environments that is passed to \eval is the environment of a
package. It refers to the \emph{namespace} of a package. A call to
\c{library(packageName)}, which loads libraries in R, for instance, will
populate environment \c{package:packageName}. It is a \emph{base} function
though, so we do not see it in the CRAN dataset. However, R still makes it
possible to access any package environments and \packageNbPackageEnvSites call
sites, \ie \packagePackageEnvSitePercent access a package environment, from
\packageNbPackageEnvPackages different packages. Those packages perform
introspection, to represent and search any environment by name, in package
\emph{envnames}, to containerize tests, as with \emph{testthat} and
\emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create
poor man's hidden functions in \emph{USSCXenaTools}. It is also used to
implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.

@@

We looked at the environments passed to all the \eval calls in our corpus.
Table~\ref{tab:environments} summarizes the results. A numeric environment class
\c{n} denotes the environment of the $n$-th call stack frame from the current
function: \c{0} is the stack of the current function, \c{1} is the parent frame. \c{global} denotes the top-level environment and \c{list} denotes a
list passed for evaluation of formulas. Environment classes of the form $n+$
denote the $n$-th environment extended with a new environment. The new
environment provides a limited form of sandboxing. All assignments using the
$\leftarrow$ function occur inside it and prevent the extended environment from
mutation. However, it is still possible to mutate the extended environment using
the $\leftarrow$ or \c{assign} functions; but, that happens rarely.

% Updated with the last run
% I made some changes and we need another run to get the right results
% The difference should be that there will much more package environments
% the loop category should disappear from the top results also.
\begin{table}[htbp]{ \centering
        \begin{tabular}{|c|c|}\hline
            Environment Class & Sites\\\hline
            \packageEnvClassNamesa &  \packageEnvClassSitePercenta\\
            \packageEnvClassNamesb &  \packageEnvClassSitePercentb\\
            \packageEnvClassNamesc &  \packageEnvClassSitePercentc\\
            \packageEnvClassNamesd &  \packageEnvClassSitePercentd\\
            \packageEnvClassNamese &  \packageEnvClassSitePercente\\
            \packageEnvClassNamesf &  \packageEnvClassSitePercentf\\
            \packageEnvClassNamesg &  \packageEnvClassSitePercentg\\
            \packageEnvClassNamesh &  \packageEnvClassSitePercenth\\
            \hline
    \end{tabular}} \label{tab:environments}
    \caption{Environments in terms of sites. One site can be passed different kinds of environments.}
\end{table}

Most of the calls happen in the environment surrounding \eval. This corresponds mostly to the default \c{envir} argument of \eval; indeed, \packageDefautEnvirSitesPercent of the sites have that default argument. A lot of sites also use \c{parent.frame()} to access their parent frame. It corresponds to environment class \c{1} but also to \c{global} when the function is called at the top level. Explicitly and directly referring to the global environment using \c{globalenv()} or \c{.GlobalEnv} is rare, as it happens only in \packageNbExplicitGlobalSites sites.

>>>>

The top-level environment in R is called the global environment. New
environments can be created using \c{new.env}, or \c{list2env}. They can be
provided a parent environment which becomes the enclosing scope of the new
environment. The new
environment provides a limited form of sandboxing. All assignments using the
$\leftarrow$ function occur inside it and prevent the extended environment from
mutation. However, it is still possible to mutate the extended environment using
the $\leftarrow$ or \c{assign} functions; but, that happens rarely.
% as.environment also can be used sometimes for that
% list2env creates an environment only if its argument envir is not provided

Another kind of environments that is passed to \eval is the environment of a
package or its \emph{namespace}. A call to
\c{library(packageName)}, which loads libraries in R, for instance, will
populate environment \c{package:packageName}. It is a \emph{base} function
though, so we do not see it in the CRAN dataset. However, R still makes it
possible to access any package environments and \packageNbPackageEnvSites call
sites, \ie \packagePackageEnvSitePercent access a package environment, from
\packageNbPackageEnvPackages different packages. Those packages perform
introspection, to represent and search any environment by name, in package
\emph{envnames}, to containerize tests, as with \emph{testthat} and
\emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create
poor man's hidden functions in \emph{USSCXenaTools}. It is also used to
implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.

%TODO: there is a difference between a package environment and a namespace. Shortly explain.

@@

We classify environments to \eval as follows:

\begin{itemize}
	\item function environments: the frame of the calling function or one of the parent frames (obtained using \c{parent.frame()}, \c{sys.frame()} or with an integer argument for \c{envir} for instance);
	\item list environments: a list, a data frame, a tibble, or \c{NULL} (seen as the empty list), passed to \eval;
	\item the global environment;
	\item package and namespace environments;
	\item the empty environment (created with \c{emptyenv()})
\end{itemize}
We also consider an orthogonal category: environments built with \c{new.env}. They can have as parent an environment of any of the other categories or another environment built with \c{new.env}.

As shown in Table~\ref{tab:highlevelenvironments}, most calls happen in a function environment, with \c{global} as a distant second.

\begin{table}[h]
	\centering
	\begin{tabular}{|c|c|c|}
		\hline
		Environment & \#sites & \%sites \\
		\hline
		Function environments &  \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
		Global &  \packageNbGlobalEnvSites & \packageNbGlobalEnvSitePercent \\
		List &  \packageNbEnvListSites & \packageNbEnvListSitePercent \\
		Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\
		Empty & \packageNbEmptyEnvSites & \packageNbEmptyEnvSitePercent \\
		\hline
		New & \packageNbNewEnvSites & \packageNewEnvSitePercent \\
		\hline
	\end{tabular} \label{tab:highlevelenvironments}
	\caption{Classification of environments per site.}
\end{table}

The calls happening in a function environment are mostly happening in the environment of the function calling \eval, as shown in Table%TODO

%TODO: add here table of possible frame numbers

Most of the calls happen in the environment surrounding \eval. This corresponds mostly to the default \c{envir} argument of \eval; indeed, \packageDefautEnvirSitesPercent of the sites have that default argument. A lot of sites also use \c{parent.frame()} to access their parent frame. It corresponds to environment class \c{1} but also to \c{global} when the function is called at the top level. Explicitly and directly referring to the global environment using \c{globalenv()} or \c{.GlobalEnv} is rare, as it happens only in \packageNbExplicitGlobalSites sites. Accessing further-away frames happens much more seldom.

%TODO: add here table of possible environment categories wrapped in new

% Up-to-date
\begin{table}[htbp]{ \centering
        \begin{tabular}{|c|c|}\hline
            Environment Class & Sites\\\hline
            \packageEnvClassNamesa &  \packageEnvClassSitePercenta\\
            \packageEnvClassNamesb &  \packageEnvClassSitePercentb\\
            \packageEnvClassNamesc &  \packageEnvClassSitePercentc\\
            \packageEnvClassNamesd &  \packageEnvClassSitePercentd\\
            \packageEnvClassNamese &  \packageEnvClassSitePercente\\
            \packageEnvClassNamesf &  \packageEnvClassSitePercentf\\
            \packageEnvClassNamesg &  \packageEnvClassSitePercentg\\
            \packageEnvClassNamesh &  \packageEnvClassSitePercenth\\
            \hline
    \end{tabular}} \label{tab:environments}
    \caption{Environments in terms of sites. One site can be passed different kinds of environments.}
\end{table}
